---
title: "Vaccination Rates in Schools Across the US"
subtitle: "Data taken from tidytuesday and Wall Street Journal"
author: "GGPlot Gods <br> Lindsey Weyant, Ali Raich, Aden Clemente"
institute: "Duke University"
date: "11/17/21"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(xaringanthemer)
library(sf)
library(viridis)
library(ggspatial)
library(usmap)
library(tidycensus)
measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-25/measles.csv')
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )
```

```{r load-data, include=FALSE}
measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-25/measles.csv')
```

```{r key, echo = FALSE, message = FALSE, warning = FALSE, include=FALSE}
census_api_key("abc8289fa2ba274ced76d97c7f8ee31666a2c931", overwrite = TRUE, install = TRUE)
 
if (FALSE) {
census_api_key("abc8289fa2ba274ced76d97c7f8ee31666a2c931", install = TRUE)
# First time, reload your environment so you can use the key without restarting R.
readRenviron("~/.Renviron")
# You can check it with:
Sys.getenv("CENSUS_API_KEY")
}
```

```{r style-slides, echo = FALSE}
style_xaringan(
  title_slide_background_image = "img/watercolour_sys02_img34_teacup-ocean.jpg"
)
```

```{r get-income-data, echo = FALSE, message = FALSE, warning = FALSE}
income <- get_acs(geography = "state", 
                       variables = "B07011_001",
                       year = 2018)
```


```{r get-race-data, echo = FALSE, message = FALSE, warning= FALSE}
race <- get_acs(geography = "state", 
                       variables = "B02001_002",
                       year = 2018)
```


```{r get-population-data, echo = FALSE, warning=FALSE, message=FALSE}
population <- get_acs(geography = "state", 
                       variables = "B01003_001",
                       year = 2018)
```

```{r variables, echo = FALSE}
measles <- measles %>%
  mutate(realrate = ifelse(mmr == -1, overall, 
                    ifelse(overall == -1, mmr, 
                    ifelse(overall == -1 && mmr == -1, -1,
                    overall)))) %>%
  filter(realrate != -1) %>%
  filter(!state %in% c("California", "Colorado", "Ohio")) %>%
  
  mutate(numvaxx = round(enroll*realrate*.01)) %>%
  mutate(unvaxx = enroll - numvaxx) %>%
  
  mutate(numxmed = round(enroll*.01*xmed)) %>%
  mutate(numxmed = ifelse(is.na(numxmed), 0, numxmed)) %>%
  
  mutate(numxrel = round(enroll*.01*xrel)) %>%
  mutate(numxrel = ifelse(is.na(numxrel), 0, numxrel)) %>%
  
  mutate(numxper = round(enroll*.01*xper)) %>%
  mutate(numxper = ifelse(is.na(numxper), 0, numxper)) %>%
  
  mutate(numxother = unvaxx - numxper - numxmed - numxrel) %>%
  mutate(numxother = ifelse(is.na(numxother), 0, numxother)) %>%

  mutate(statefac = as.factor(state)) %>%

  group_by(state) %>%
  mutate(statemean = mean(realrate))
```

```{r join-income-statemean, echo = FALSE}
incomerates <- left_join(measles, income, by = c("state" = "NAME"))
```


class: inverse, middle, center

# Intro

---

## Research Question

How do measles vaccination rates vary across the country in different types of school?

---

## Where did our data come from?

-tidytuesday measles vaccination rates

-Wall Street Journal article from October 2019

--

-"What's the Measles Vaccination Rate at Your Child's School?"

---

## What's important about our data?

-WHO recommends a 95% vaccination rate among elementary schools

-It's important to note that there is not a universal method for collecting and keeping track of immunization rates so each state's data set is slightly different.

---

## Data Wrangling

-Mutation for if school had mmr but not overall:

--

-If school had overall vaccination rate, we used overall, but if there was only mmr, we used the mmr rate.

--

-Majority of our analysis by type of school was conducted between private, public, and charter schools.

---

class: inverse, middle, center

#Vaccination Exemption Type Graphs

##Preliminary Analysis

---

```{r vaxgraph, echo = FALSE, warning = FALSE}
measles %>%
  group_by(type) %>%
  mutate(averagevax = mean(realrate)) %>%
  select(type, averagevax) %>%
  distinct() %>%
  ggplot(aes(x = type, y = averagevax)) +
  geom_col() +
  labs(title = "Measles Vaccination Rates Across Different School Types", x = "School Type", y = "Pecentage of Students Vaccinated")
```
---

```{r religiousexemption, echo = FALSE, warning = FALSE}
measles %>%
  mutate(xrel = ifelse(is.na(xrel), 0, xrel)) %>%
  group_by(type) %>%
  mutate(averagerel = mean(xrel)) %>%
  select(type, averagerel) %>%
  distinct() %>%
  ggplot(aes(x = type, y = averagerel)) +
  geom_col() +
  labs(title = "Religious Exemption Rates Across Different School Types", 
       x = "School Type", y = "Pecentage of Students with Religious Exemption")
```

---

```{r personalexemption, echo = FALSE}
measles %>%
  mutate(xper = ifelse(is.na(xper), 0, xper)) %>%
  group_by(type) %>%
  mutate(averageper = mean(xper)) %>%
  select(type, averageper) %>%
  distinct() %>%
  ggplot(aes(x = type, y = averageper)) +
  geom_col() +
  labs(title = "Personal Exemption Rates Across Different School Types", 
       x = "School Type", y = "Pecentage of Students with Personal Exemption")
```

---

```{r medicalexemption, echo = FALSE}
measles %>%
  mutate(xmed = ifelse(is.na(xmed), 0, xmed)) %>%
  group_by(type) %>%
  mutate(averagemed = mean(xmed)) %>%
  select(type, averagemed) %>%
  distinct() %>%
  ggplot(aes(x = type, y = averagemed)) +
  geom_col() +
  labs(title = "Medical Exemption Rates Across Different School Types", 
       x = "School Type", y = "Pecentage of Students with Medical Exemption")
```

---

```{r vaxbystatemap, echo=FALSE}
plotdata <- measles %>%
filter(realrate != (-1)) %>%
group_by(state) %>%
summarise(statemean = mean(realrate))
plot_usmap(data=plotdata, values = "statemean") +
  labs(title = "Vaccination Rate by State", fill = "Vaccination Rate") +
  theme(panel.background = element_rect(color = "black", fill = "lightblue"))
```

---

```{r income-graph, echo = FALSE, fig.width=6, fig.height=4}
income %>%
  filter(NAME %in% c("Arizona", "Arkansas", "Connecticut", "Florida", "Idaho", "Illinois", "Iowa", "Maine", "Massachusetts", "Michigan", "Minnesota", "Missouri", "Montana", "New Jersey", "New York", "North Carolina", "North Dakota", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "Wisconsin")) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_point() +
  scale_y_discrete() +
  labs(x = "Median Income", y = "State", title = "Median Income by State")
```

---

class: inverse, middle, center

#T Tests and Anova

---

##T Tests

```{r vaxbytypefilter, echo=FALSE}
measles3 <- measles %>%
 filter(type %in% c("Public", "Private"))

measles4 <- measles %>%
  filter(type %in% c("Public", "Charter"))

measles5 <- measles %>%
  filter(type %in% c("Private", "Charter")) 
  
```

```{r vaxbytype-ttest, echo=FALSE}
t.test(measles3$realrate~measles3$type) %>%
  tidy() %>%
  knitr::kable(caption="Output for T-Test between Public and Private Schools", format="html")

t.test(measles4$realrate~measles4$type) %>%
  tidy() %>%
  knitr::kable(caption="Output for T-Test between Public and Charter Schools", format= "html")

t.test(measles5$realrate~measles5$type) %>%
  tidy() %>%
  knitr::kable(caption="Output for T-Test between Private and Charter Schools", format = "html")

```
---

## ANOVA

```{r vaxbystateanova, echo=FALSE}
tidy(aov(realrate~state,data=measles)) %>%
  knitr::kable(caption="ANOVA")
```

---

class: inverse, middle, center

# Regression Analysis

---

##Logistic Regression

```{r logistic-reg, echo = FALSE, fig.width=6, fig.height=4}
measles$statefac = relevel(measles$statefac, ref = "Arkansas")

measlereg <- glm(cbind(numvaxx, unvaxx) ~ statefac, data=measles, family = binomial)
tidy(measlereg, exponentiate = TRUE) %>%
    knitr::kable(caption="Output for Logistic Regression")
```

---

##Linear Regression

```{r linreg-vaxbyincome, echo = FALSE}
linear_reg() %>%
 set_engine("lm") %>%
  fit(statemean ~ estimate, data = incomerates) %>%
  tidy() %>%
  knitr::kable(caption="Output for Linear Regression with Respect to Income")
```

---

## Linear Regression Graph

```{r linreg-vaxbyincomegraph, echo=FALSE}
ggplot(incomerates, aes(x = estimate, y=statemean)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
labs(x = "Average State Income", y="State Mean Vax Rate", title = "State Mean Vax Rate by Average State Income")
```
---

## Linear Regression Equation

Equation for predicting state mean vax rate: $\hat{y} = 81.60 + 0.0004092 * x_i$

---

## Summary

-There is a relationship between overall vaccination rate and type of school

--

-Charter schools had the lowest vaccination rate, public schools had the highest

--

-Vaccination rate and state

-Vaccination rate and median state income

---

class: inverse, middle, center

# Further research

